---
title: "Communicating Statistical Results"
format: 
  revealjs:
    footer: "[Teaching Intermediate R](https://github.com/atheobold/uscots-intermediate-r)"
    theme: [simple, styles.scss]
    embed-resources: true
editor: source
---

```{r}
#| include: false
library(countdown)
```

## The "90/10" rule

> 90% of the work you do will occupy only 10% of your presentation time/space.

* Students tend to make reports and dashboards using **the timeline they experienced** in the analysis

* Instead, focus on **conclusions and takeaways**

* Differentiate between **communicating results** and **documenting a process** (both are important!)

## The "90/10" rule

10%:

* Data **cleaning and wrangling** details

* **Assumption checking** for models or statistical tests

* **Combined results** all in one visual

90%:

* Specific **conclusions** from tests - "we found evidence that..."

* Plots of **model predictions** and **summary statistics**

* Visuals highlighting **individual takeaways** - "California has higher predicted prices than the Midwest"


## The "Newspaper Headline" approach

> For each analysis, what is the headline a newspaper would use.

* **NO:** Statements of general accomplishment - "We fit a model to predict prices of fast food items."

* **NO:** Model assessment - "Our model has 93% accuracy on a test set."

* **NO:** Results in technical form - "The variable found to be most significant was location of store."

* **NO:** Unspecific results - "Location of fast food store impacts price."

* **YES:** Results *of public interest* in *plain terms* with *specific knowledge*: "Fast food prices are 50% higher in California than in the Midwest." 

## YOUR TURN

::: {.r-fit-text}
Brainstorm a list of headlines that could be produced from your data.

Make these specific, not meta, even though the numbers will be made up - e.g. "Fast food prices are 50% higher in California than in the Midwest." not "Find the differences in price by region".
:::

```{r}
#| echo: false
countdown(minutes = 3)
```


## Dashboards as "Q&A"

> The goal of an **interactive** dashboard is to **replace you**, by giving the audience a way to ask and answer questions themselves.

* Anticipate **follow up questions** to the main "headlines".

* Go from **general** to **specific** - e.g. "Here is a plot of the average price of burgers over time.  Use this drop-down to filter by state."

* Keep your dashboard **sanitized** against bad statistical choices - e.g., don't allow users to select "Show logistic regression" for numeric data.

## "Why should I care?"

> Dashboards and presentations are for your target audience, not your peers.

* What is interesting to a **data scientist** might not be interesting to your **target audience**.

* Use **personas**: 
    + Give your client a name
    + Why does this client want your analysis?
    + What is their specific stake in the data?
    + What level of technical content will they understand and appreciate?
    
* Stay focused on how the client **will use the analysis** to make decisions.

## Example

> My client is Nisha, a CEO of a small restaurant chain that competes with local fast food.  She has a degree in business administration, so she understands basic statistics. She has hired me to analyze price trends in the food industry.

* Nisha is a CEO of a small business - she is probably interested in *how to keep her restaurant solvent*

* We can quote *basic statistics*, but we should not go deep into modeling math.

* She has asked specifically about *prices*, so our presentation should focus on that.

* She has asked about *trends*, so our "newspaper headline" should be about changes over time.

* We should state how our analysis will *help her set her prices* at her restaurants.

## YOUR TURN

::: {.r-fit-text}
Come up with **three different personas** that your students might target with their dashboards.

How would you expect the dashboard to be **different** between these three personas?
:::

```{r}
#| echo: false
countdown(minutes = 3)
```

## Activity ideas

**Can the data answer this question?**

* Give students a dataset and a list of questions

**Is this conclusion supported?**

* Give a *visualization* or *test result*, and a client quote - "This means I should sell more hamburgers!"

* Ask students if they think the result supports the conclusion.

**"Client" feedback**

* Invent a mock client for *all* student groups to target

* Have student groups swap projects and imagine themselves as the clients


## Create your Project B Dashboard Assignment

Things to include:

* Who is the **target audience** that the student should address?

* Will you require any **specific research questions** to be addressed, or is it open-ended?

* Will you require any **specific dashboard elements**, such as interactivity?

* How will you grade:
    + the **design** elements?
    + the **technical** elements?
    + the **communication** elements?






